---
title: "Appendix"
author: "Yixuan Jiao, Hongpu Min, Qingzhen Sun, Yining Chen"
date: "2023-12-10"
output: pdf_document
---

```{r message= FALSE}
library(tidyverse)
library(survival)
library(survminer)
library(biostat3)
library(finalfit)
library(dplyr)
library(kableExtra)
library(ggplot2)
library(ggpubr)
library(riskRegression)
library(regclass)
```

# EDA
```{r}
aids <- read.csv('data/AIDS_Clinical_Trials_Group175.csv')
aids <- aids%>%
  mutate(cid = as.factor(cid),
         
         trt = as.factor(trt),treat=as.factor(treat),
                                                                     hemo = as.factor(hemo),
                                                                     homo = as.factor(homo),
                                                                     drugs = as.factor(drugs),
                                                                     race = as.factor(race),
                                                                     gender = as.factor(gender),
                                                                     str2 = as.factor(str2),
                                                                     symptom = as.factor(symptom))
```

```{r}
aids <- aids%>%
  mutate(cid=factor(cid,
  levels = c(0,1),
  labels = c("Censoring", 
             "Failure")),
  treat=factor(treat,
  levels = c(0,1),
  labels = c("ZDV only", 
             "Others")),
  hemo=factor(hemo,
  levels = c(0,1),
  labels = c("No", 
             "Yes")),
  homo=factor(homo,
  levels = c(0,1),
  labels = c("No", 
             "Yes")),
  gender=factor(gender,
  levels = c(0,1),
  labels = c("Female", 
             "Male")),
  race=factor(race,
  levels = c(0,1),
  labels = c("White", 
             "Non-white")),
  drugs=factor(drugs,
  levels = c(0,1),
  labels = c("No", 
             "Yes")),
  symptom=factor(symptom,
  levels = c(0,1),
  labels = c("No", 
             "Yes"))
  )
```

## sex age race hemo homo drug kanor symptom cd40
```{r}
explanatory = c("age", "hemo", "homo", "race","gender","drugs","karnof","cd40","symptom","cd80","wtkg")
dependent = "treat"
baseline <- aids %>%
  mutate(
        cd80 = ff_label(cd80, "CD8 Count"),
        wtkg = ff_label(wtkg, "Weight"),
        gender = ff_label(gender, "Gender"),
        hemo = ff_label(hemo, "Hemophilia"),
        homo = ff_label(homo, "Homosexuality"),
        race = ff_label(race, "Race"),
        drugs = ff_label(drugs, "History of IV drug use "),
        karnof = ff_label(karnof, "Karnofsky score of 100"),
        cd40 = ff_label(cd40, "CD4 count"),
        age = ff_label(age, "Age"),
        symptom = ff_label(symptom, "Symptomatic infection"),
        treat = ff_label(treat, "Treatment")
    )%>%
  summary_factorlist(dependent, explanatory,column = TRUE, total_col = TRUE,,col_totals_prefix = "N=",add_dependent_label = TRUE) 

baseline

kable(baseline,caption = "Base-Line Characteristics of the Patients According to the Treatment Indicator", booktabs=T)%>% kable_styling(latex_option = c("hold_position"), position = "center")
```
```{r}
ggplot(data=aids, aes(x=treat, fill=cid)) +
  geom_bar()+
  scale_fill_brewer(palette="Pastel1")+
  ggtitle("Figure 1: Distribution of  Treatment VS. Patient Status")+
  guides(fill = guide_legend(title = "Treatment"))
```

```{r}
ggplot(aids, aes(x=age, fill=cid)) +
  geom_histogram(alpha=0.5, position="identity")+
  scale_fill_brewer(palette ="Pastel1")+
  ggtitle("Figure 2: Distribution of Patient Age VS. Status")

ggplot(aids, aes(x=wtkg, fill=cid)) +
  geom_histogram(alpha=0.5, position="identity")+
  scale_fill_brewer(palette ="Pastel1")+
  ggtitle("Figure 3: Distribution of Patient Weight VS. Status")
```
```{r}
ggplot(aids, aes(x=time, fill=cid)) +
  geom_histogram(alpha=0.5, position="identity")+
  scale_fill_brewer(palette ="Pastel1")+
  ggtitle("Figure 4: Distribution of Time to Failure or Censoring")
```
```{r}
p1 <- ggplot(aids,aes(x= treat, y=cd40,color=treat)) + geom_boxplot(show.legend = FALSE)+labs(x="")+scale_color_manual( values = c("#DC9FB4", "mediumpurple"))+theme(axis.title.y = element_text(size=8))+
  theme_minimal()+ 
 scale_y_continuous(limits = c(0, 1000))


p2 <- ggplot(aids,aes(x= treat, y=cd420,color=treat)) + geom_boxplot(show.legend = FALSE)+labs(x="")+scale_color_manual( values = c("#DC9FB4", "mediumpurple"))+theme(axis.title.y = element_text(size=8))+
  theme_minimal()+
 scale_y_continuous(limits = c(0, 1000))

p <- ggpubr::ggarrange(p1, p2, ncol=2,nrow = 1,common.legend = TRUE)

ggpubr::annotate_figure(p, top = ggpubr::text_grob("Figure 5: CD4 Count Change VS. Treatment Groups", color = "darkgrey",face = "bold", size = 13))
```

```{r}
p3 <- ggplot(aids,aes(x= treat, y=cd80,color=treat)) + geom_boxplot(show.legend = FALSE)+labs(x="")+scale_color_manual( values = c("#DC9FB4", "mediumpurple"))+theme(axis.title.y = element_text(size=8))+
  theme_minimal()+ 
 scale_y_continuous(limits = c(0, 4000))

p4 <- ggplot(aids,aes(x= treat, y=cd820,color=treat)) + geom_boxplot(show.legend = FALSE)+labs(x="")+scale_color_manual( values = c("#DC9FB4", "mediumpurple"))+theme(axis.title.y = element_text(size=8))+
  theme_minimal()+ 
 scale_y_continuous(limits = c(0, 4000))

p<- ggarrange(p3,p4,ncol=2,align = "hv",nrow = 1,common.legend = TRUE)
annotate_figure(p, top = text_grob("Figure 6: CD8 Count Change VS. Treatment Groups", color = "#0F2540",face = "bold", size = 13))
```


# Non-parametric tests
```{r}
aids <- read_csv("data/AIDS_Clinical_Trials_Group175.csv")
aids %>% head()
```

```{r}
#KM estimate 
km_fit_trt <- survfit(Surv(time, cid) ~ trt, data = aids)
pkm <- km_fit_trt %>% ggsurvplot(data = aids,
                      fun = "pct", #can be replaced by cum hazard
                      conf.int = TRUE,
                      risk.table = TRUE,
                      pval = TRUE,  
                      pval.coord = c(0,55),
                      fontsize = 2.5,
                      ggtheme = theme_light(),
                      xlab =  "Time in days",
                      title = "Figure 4: Kaplan-Meier Survival Function Estimate",
                      legend.title = "",
                      legend.labs = c("ZDV only","ZDV + ddl", "ZDV + Zal", "ddl only"),
                      ylim = c(55, 100))

hkm <- km_fit_trt %>% ggsurvplot(data = aids,
                      fun = "cumhaz", #can be replaced by cum hazard
                      conf.int = TRUE,
                      risk.table = TRUE,
                      pval = TRUE,  
                      fontsize = 2.5,
                      ggtheme = theme_light(),
                      xlab =  "Time in days",
                      title = "Figure 5: Kaplan-Meier Cumulative Hazard Function",
                      legend.title = "",
                      legend.labs = c("ZDV only","ZDV + ddl", "ZDV + Zal", "ddl only"))
pkm
hkm
```


```{r}
quantile(km_fit_trt,probs = c(0.1,0.15,0.2))$quantile %>% knitr::kable(caption = "t1")
```
Table: Quantile Survival Time (in days) by Treatments

|      Treatment|  $90^{th}$|  $85^{th}$|  $80^{th}$|
|:-----|---:|---:|---:|
|ZDV only | 347 (284,406)| 468 (394,557)| 569 (484,649)|
|ZDV + ddl | 626 (559,721)| 822 (691,929)| 986 (876,NA)|
|ZDV + Zal | 610 (476,748)| 806 (720,910| 972 (867,NA)|
|ddl only | 537 (476,631)| 672 (613,813)| 898 (760,994)|

```{r}
survdiff(Surv(time, cid) ~ trt, data = aids)
```

```{r}
aids %>% 
  mutate(trt = case_when(trt == "0" ~ "ZDV only",
                          trt == "1" ~ "ZDV + ddl",
                          trt == "2" ~ "ZDV + Zal",
                          trt == "3" ~ "ddl only")) %>%
  pairwise_survdiff(Surv(time, cid) ~ trt, data = .,p.adjust.method = "BH") %>%
  broom::tidy()
```
Table: Pairwise Log-rank Tests by Treatments

|group1    |group2    |   P-value|
|:---------|:---------|---------:|
|ZDV + ddl |ddl only  | 0.2784441|
|ZDV + Zal |ddl only  | 0.4776330|
|ZDV + Zal |ZDV + ddl | 0.6362565|
|ZDV only  |ddl only  | 0.0000075$^{***}$|
|ZDV only  |ZDV + ddl | 0.0000000$^{***}$|
|ZDV only  |ZDV + Zal | 0.0000002$^{***}$|


### KM curve for zdv history stratify
```{r}
# non-ZDV treatment history
aids_non_zdv <- aids %>%
  filter(oprior == 1)
```

```{r}
km_fit_trt1 <- survfit(Surv(time, cid) ~ trt, data = aids_non_zdv)
km_fit_trt1 %>% ggsurvplot(data = aids_non_zdv,
                      fun = "pct", 
                      conf.int = TRUE,
                      risk.table = TRUE,
                      fontsize = 2,
                      ggtheme = theme_light(),
                      title = "Kaplan-Meier Estimate Without ZDV History",
                      legend.title = "",
                      legend.labs = c("ZDV only","ZDV + ddl", "ZDV + Zal", "ddl only"),
                      ylim = c(55, 100))
```

```{r}
# ZDV in the 30 days prior to 175
aids_z30 = aids%>%
  filter(z30 == 1)
```

```{r}
km_fit_trt2 <- survfit(Surv(time, cid) ~ trt, data = aids_z30)
km_fit_trt2 %>% ggsurvplot(data = aids_z30,
                      fun = "pct", 
                      conf.int = TRUE,
                      risk.table = TRUE,
                      fontsize = 2,
                      ggtheme = theme_light(),
                      title = "Kaplan-Meier Estimate With ZDV 30 days prior",
                      legend.title = "",
                      legend.labs = c("ZDV only","ZDV + ddl", "ZDV + Zal", "ddl only"),
                      ylim = c(55, 100))
```

```{r}
# ZDV prior to 175
aids_zprior = aids%>%
  filter(zprior == 1)
```

```{r}
km_fit_trt3 <- survfit(Surv(time, cid) ~ trt, data = aids_zprior)
km_fit_trt3 %>% ggsurvplot(data = aids_zprior,
                      fun = "pct", 
                      conf.int = TRUE,
                      risk.table = TRUE,
                      fontsize = 2,
                      ggtheme = theme_light(),
                      title = "Kaplan-Meier Estimate With ZDV 175 days prior to study",
                      legend.title = "",
                      legend.labs = c("ZDV only","ZDV + ddl", "ZDV + Zal", "ddl only"),
                      ylim = c(55, 100))
```

```{r}
survfit_result_zdv <- survfit(Surv(time, cid) ~ strata(oprior) + trt, data = aids)
survfit_result_zdv
```


### KM Curve for medical history stratify
```{r}
# patient with drug used and has hemophilia
aids_hemo_drug = aids%>%
  filter(hemo==1)%>%
  filter(drugs ==1)
```

```{r}
km_fit_trt4 <- survfit(Surv(time, cid) ~ trt, data = aids_hemo_drug)
km_fit_trt4 %>% ggsurvplot(data = aids_hemo_drug,
                      fun = "pct", 
                      conf.int = TRUE,
                      risk.table = TRUE,
                      fontsize = 2,
                      ggtheme = theme_light(),
                      title = "Kaplan-Meier Estimate With Drugs and Hemophilia",
                      legend.title = "",
                      legend.labs = levels(aids$trt),
                      ylim = c(55, 100))
```

```{r}
# patient with drug used and doesn't have hemophilia
aids_drug = aids%>%
  filter(hemo==0)%>%
  filter(drugs ==1)
```

```{r}
km_fit_trt5 <- survfit(Surv(time, cid) ~ trt, data = aids_drug)
km_fit_trt5 %>% ggsurvplot(data = aids_drug,
                      fun = "pct", 
                      conf.int = TRUE,
                      risk.table = TRUE,
                      fontsize = 2,
                      ggtheme = theme_light(),
                      title = "Group 2: Kaplan-Meier Estimate With Drug Used",
                      legend.title = "",
                      legend.labs = c("ZDV only","ZDV + ddl", "ZDV + Zal", "ddl only"),
                      ylim = c(55, 100))
```

```{r}
# patient without drug used but has hemophilia
aids_hemo = aids%>%
  filter(hemo==1)%>%
  filter(drugs ==0)
```

```{r}
km_fit_trt6 <- survfit(Surv(time, cid) ~ trt, data = aids_hemo)
km_fit_trt6 %>% ggsurvplot(data = aids_hemo,
                      fun = "pct", 
                      conf.int = TRUE,
                      risk.table = TRUE,
                      fontsize = 2,
                      ggtheme = theme_light(),
                      title = "Group 3: Kaplan-Meier Estimate With Hemophilia",
                      legend.title = "",
                      legend.labs = c("ZDV only","ZDV + ddl", "ZDV + Zal", "ddl only"),
                      ylim = c(55, 100))
```

```{r}
# patient without drug used and doesn't have hemophilia
aids_nomedical = aids%>%
  filter(hemo==0)%>%
  filter(drugs ==0)
```

```{r}
km_fit_trt7 <- survfit(Surv(time, cid) ~ trt, data = aids_nomedical)
km_fit_trt7 %>% ggsurvplot(data = aids_nomedical,
                      fun = "pct", 
                      conf.int = TRUE,
                      risk.table = TRUE,
                      fontsize = 2,
                      ggtheme = theme_light(),
                      title = "Group 4:Kaplan-Meier Estimate Without medical history",
                      legend.title = "",
                      legend.labs = c("ZDV only","ZDV + ddl", "ZDV + Zal", "ddl only"),
                      ylim = c(55, 100))

```

```{r}
survfit_result_drug <- survfit(Surv(time, cid) ~ strata(drugs) + trt, data = aids)
survfit_result_drug

survfit_result_hemo <- survfit(Surv(time, cid) ~ strata(hemo) + trt, data = aids)
survfit_result_hemo
```


### KM curve for patient with or without azt therapy before

```{r}
# patient with azt theorpy before
aids_azt_yes = aids%>%
  filter(str2==1)
```

```{r}
km_fit_trt8 <- survfit(Surv(time, cid) ~ trt, data = aids_azt_yes)
km_fit_trt8 %>% ggsurvplot(data = aids_azt_yes,
                      fun = "pct", 
                      conf.int = TRUE,
                      risk.table = TRUE,
                      fontsize = 2,
                      ggtheme = theme_light(),
                      title = "Kaplan-Meier Estimate Without medical history",
                      legend.title = "",
                      legend.labs = levels(aids$trt),
                      ylim = c(55, 100))

```


```{r}
# patient without azt therapy before
aids_azt_no <- 
  aids %>%
  filter(str2==0)
```

```{r}
km_fit_trt9 <- survfit(Surv(time, cid) ~ trt, data = aids_azt_no)
km_fit_trt9 %>% ggsurvplot(data = aids_azt_no,
                      fun = "pct", 
                      conf.int = TRUE,
                      risk.table = TRUE,
                      fontsize = 2,
                      ggtheme = theme_light(),
                      title = "Kaplan-Meier Estimate Without medical history",
                      legend.title = "",
                      legend.labs = levels(aids$trt),
                      ylim = c(55, 100))

```

```{r}
survfit_result_azt <- survfit(Surv(time, cid) ~ strata(str2) + trt, data = aids)
survfit_result_azt
```

# Cox-PH model

```{r}
aids = read_csv("data/AIDS_Clinical_Trials_Group175.csv") %>% mutate(trt = as.factor(trt),
                                                                     hemo = as.factor(hemo),
                                                                     homo = as.factor(homo),
                                                                     drugs = as.factor(drugs),
                                                                     race = as.factor(race),
                                                                     gender = as.factor(gender),
                                                                     str2 = as.factor(str2),
                                                                     symptom = as.factor(symptom)) %>% dplyr::select(time, cid, trt, age, wtkg, hemo, homo, drugs, karnof, race, gender, str2, symptom, cd40, cd80) %>% arrange(time) %>% drop_na()
```

#### model selection

*selection on personal information*
```{r}
selectCox(Surv(time, cid) ~ trt + age + wtkg + homo + race + gender, data = aids, rule = "aic")
```
$In
[1] "trt" "age"

*selection on medical history/treatment history/lab results*
```{r}
selectCox(Surv(time, cid) ~ trt + hemo + drugs + karnof + str2 + symptom + cd40 + cd80, data = aids, rule = "aic")
```
$In
[1] "trt"     "drugs"   "karnof"  "str2"    "symptom" "cd40"    "cd80"  

*selection with interaction*
```{r}
selectCox(Surv(time, cid) ~ trt + drugs + karnof + str2 + symptom + cd40 + cd80 + age + age * drugs + age * trt, data = aids, rule = "aic")
```
$In
[1] "trt"         "karnof"      "str2"        "symptom"     "cd40"        "cd80"        "drugs * age"

*final model*
```{r}
cox1 = coxph(Surv(time, cid) ~ trt + karnof + str2 + symptom + cd40 + cd80 + age * drugs, data = aids)
```

```{r}
summary(cox1)
```

*final model with age_group*
```{r}
aids = aids %>% mutate(age_group = as.factor(ifelse(age >= 11 & age <= 30, "11-30", ifelse(age >= 31 & age <= 50, "31-50", "51-70"))))
```

```{r}
cox2 = coxph(Surv(time, cid) ~ trt + age_group + drugs + karnof + str2 + symptom + cd40 + cd80 + age_group * drugs, data = aids)
```

```{r}
summary(cox2)
```

#### model checking

*check multicollinearity*
```{r}
VIF(cox2)
```

*Plot log-log survival curve*
```{r}
par(mfrow = c(2,4))
var_list = names(aids)

for (i in var_list) {
plot(survfit(Surv(time, cid) ~ aids[[i]], data = aids),
fun = 'cloglog',
conf.int = FALSE,
col = 1,
lty = 1,
xlab = "Time",
ylab = "Log-log of S(t)",
main = "Log-Log Survival Curves")
}
```

*Plot the observed and fitted*
```{r}
par(mfrow = c(1,1))

plot(survfit(Surv(time, cid) ~ 1, data = aids),
conf.int = FALSE,
col = 1,
lty = 1,
ylim = c(0.55,1),
xlab = "Time",
ylab = "Survival Probability",
main = "Observed vs Fitted Survival Curves")

lines(survfit(cox2),
conf.int = FALSE,
col = 2,
lty = 2,
ylim = c(0.55,1))

legend("bottomleft",
legend = c("KM", "Cox"),
col = 1:2,
lty = c(1, 2),
cex = 0.5,
merge = TRUE)
```

```{r}
broom::tidy(cox2) %>% 
  mutate(`exp(estimate)` = exp(estimate)) %>% 
  relocate(`exp(estimate)`, .after = estimate) #%>%
  #kable(caption = "Summary of Final Cox-PH model")
```

Table: Summary of Final Cox-PH model

|term                  |   estimate| exp(estimate)| std.error|  statistic|   p.value|
|:---------------------|----------:|-------------:|---------:|----------:|---------:|
|trt1                  | -0.8001201|     0.4492750| 0.1242171| -6.4413036| <0.0000001|
|trt2                  | -0.6657643|     0.5138806| 0.1216476| -5.4728949| <0.0000001|
|trt3                  | -0.5673093|     0.5670492| 0.1159872| -4.8911374| 0.0000010|
|age_group31-50        | -0.0962569|     0.9082307| 0.1037108| -0.9281280| 0.3533412|
|age_group51-70        |  0.3463719|     1.4139283| 0.1901102|  1.8219527| 0.0684622|
|drugs1                | -0.3664675|     0.6931787| 0.3445059| -1.0637481| 0.2874428|
|karnof                | -0.0238122|     0.9764691| 0.0070067| -3.3984920| 0.0006776|
|str21                 |  0.3772408|     1.4582554| 0.0959051|  3.9334819| 0.0000837|
|symptom1              |  0.4363595|     1.5470649| 0.1030943|  4.2326242| 0.0000231|
|cd40                  | -0.0041861|     0.9958226| 0.0004476| -9.3526642| <0.0000001|
|cd80                  |  0.0004689|     1.0004690| 0.0000830|  5.6524859| <0.0000001|
|age_group31-50:drugs1 |  0.0704626|     1.0730045| 0.3831692|  0.1838943| 0.8540964|
|age_group51-70:drugs1 |  0.4512700|     1.5703052| 0.6975511|  0.6469347| 0.5176742|